---
title: "DATA607_Final_Project"
author: "Wilson Ng & Neil Hodgkinson"
date: "2022-11-25"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mongolite)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(keyring)
library(readr)
library(stringr)
```

## 1.Connecting to MongoDB to access Dr.Semmelweis's data on childbirth deaths.

```{r}
# making rstudio ask me for the password since I don't want to expose my pw when I upload this file to GitHub.

mongo_server <- 'mongodb+srv://'
username <- 'wilsonnrdb:'
password <- .rs.askForPassword("Database Password:")
cluster <- '@cluster0.cyjo0wo.mongodb.net/'

connection_string = paste(mongo_server, username, password, cluster, sep = '')

# Two collections - one has yearly data and the other one has monthly data.
yearly_collection = mongo(collection='yearly', db='DATA607', url=connection_string)
monthly_collection = mongo(collection= 'monthly', db='DATA607', url=connection_string)
```

## 2.Examining the yearly data

```{r}
yearly_query = yearly_collection$find()

yearly_query
```
## 3.Examining the monthly data

```{r}
monthly_query = monthly_collection$find()

monthly_query
```
## 4.Calculating the porportion of deaths out of the number of women giving birth.

The table above shows the number of women giving birth at the two clinics at the Vienna General Hospital for the years 1841 to 1846.  

```{r}
# Adding a new column to yearly with proportion of deaths per no. births
yearly_query <- yearly_query %>% mutate(proportion_deaths = deaths / births)

# Converting Year column's data type to double.
yearly_query$year <- as.numeric(yearly_query$year)

# Print out yearly
yearly_query
```
## 5. Plotting the proportion of deaths at both clinic 1 and clinic 2.

```{r}
# Setting the size of plots in this notebook
options(repr.plot.width=7, repr.plot.height=4)

# Plot yearly proportion of deaths at the two clinics
ggplot(yearly_query, aes(year, proportion_deaths, color = clinic)) +
    geom_line()
```
## 6. Examining the monthly data from Clinic 1.

```{r}
# Adding a new column with proportion of deaths per no. births
monthly_query <- monthly_query %>% mutate(
    proportion_deaths = deaths / births
    )

# Converting Date column's chr data type into date data type
monthly_query$date <- ymd(monthly_query$date)

# Print out the first rows in monthly
head(monthly_query)
```
```{r}
# Plot monthly proportion of deaths to see if there's any obvious decline in deaths.
ggplot(monthly_query, aes(date, proportion_deaths)) +
    geom_line() +
    labs(x = "Date", y = "Proportion of Deaths", title = "Monthly Proportion of Deaths")
```
## 7. The effect of handwashing highlighted

Dr.Semmelweis made handwashing obligatory in the summer of 1847.
The effect of handwashing is made even more clear if we highlight this in the graph.

```{r}
# From this date handwashing was made mandatory
handwashing_start = as.Date('1847-06-01')

# Add a TRUE/FALSE column to monthly called handwashing_started
monthly_query <- monthly_query %>% mutate(
    handwashing_started = date >= handwashing_start
    )

# Plot monthly proportion of deaths before and after handwashing
ggplot(monthly_query, aes(date, proportion_deaths, color = handwashing_started)) +
    geom_line() +
    labs(x = "Date", y = "Proportion of Deaths", title = "Monthly Proportion of Deaths")
```
## 8.How much did it reduce the monthly proportion of deaths on average?

It reduced the proportion of deaths by around 8 percentage points.

```{r}
# Calculating the mean proportion of deaths 
# before and after handwashing.

monthly_summary <- monthly_query %>% 
    group_by(handwashing_started) %>%
    summarise(mean_proportion_deaths = mean(proportion_deaths))

# Printing out the summary.
monthly_summary
```
## A statistical analysis of Semmelweis handwashing data

Proportion of deaths increased by between 6.7 and 10 percentage points when doctors didn't wash their hands, according to a 95% confidence interval.

```{r}
# Calculating a 95% Confidence intrerval using t.test 
test_result <- t.test( proportion_deaths ~ handwashing_started, data = monthly_query)
test_result
```

## UNICEF data on childbirth deaths

```{r}
urlfile="https://raw.githubusercontent.com/Nhodgkinson/607-Final-Project/main/JMP-WASH-in-health-care-facilities-2022-data-by-country(2).csv"

unicef_df_all<-read_csv(url(urlfile))
unicef_df_all$...3 <- str_replace_all(unicef_df_all$...3 , " ", "")

#Cleaning data frame from wide to long. Removing unneccasary columns and focusing in on Hospital data

hos_unicef_df<-unicef_df_all[,c(1:4,20:24)]
names(hos_unicef_df) <- hos_unicef_df[1,]
hos_unicef_df<-hos_unicef_df[-1,]
hos_unicef_df$`Population(thousands)`<-as.numeric(hos_unicef_df$`Population(thousands)`)
hos_unicef_df$`Population(thousands)`<-hos_unicef_df$`Population(thousands)`*(1000)
```

```{r}
urlfile2="https://raw.githubusercontent.com/Nhodgkinson/607-Final-Project/main/API_SP.DYN.IMRT.IN_DS2_en_csv_v2_4701014.csv"

#Cleaning data frame from wide to long. Only looking at years 2000-2005

infant_mortality_df<-read_csv(url(urlfile2))
infant_mortality_df<-infant_mortality_df[,c(1,5:50)]
infant_mortality_df<-gather(infant_mortality_df, "Year", "Infant Mortatlity per 1,000 Births", 2:47)
infant_mortality_df$Year<-as.numeric(infant_mortality_df$Year)
infant_mortality_df<-infant_mortality_df %>%
  subset(Year > 1999 & Year < 2006)
```

```{r}
#Increase in handwashing facilities

increase_hw_df<-hos_unicef_df %>%
  subset(Year > 1999 & Year < 2006)%>%
  filter(`COUNTRY, AREA OR TERRITORY` == "Ghana"|`COUNTRY, AREA OR TERRITORY` == "Rwanda")
increase_hw_df<-increase_hw_df[,c(1,2,8)]
```

```{r}
#Decrease in handwashing facilities

decrease_hw_df<-hos_unicef_df %>%
  subset(Year > 1999 & Year < 2006)%>%
  filter(`COUNTRY, AREA OR TERRITORY` == "Zambia"|`COUNTRY, AREA OR TERRITORY` == "Kenya")
decrease_hw_df<-decrease_hw_df[,c(1,2,8)]
```

```{r}
#no change in number of handwashing facilities

same_hw_df<-hos_unicef_df %>%
  subset(Year > 1999 & Year < 2006)%>%
  filter(`COUNTRY, AREA OR TERRITORY` == "Myanmar"|`COUNTRY, AREA OR TERRITORY` == "Mauritania")
same_hw_df<-same_hw_df[,c(1,2,8)]
```

```{r}
#Joining hospital df and mortality df

increase_merge_df<-merge(x=increase_hw_df,y=infant_mortality_df,by= 1 ,all.x=TRUE)
same_merge_df<-merge(x=same_hw_df,y=infant_mortality_df,by= 1 ,all.x=TRUE)
decrease_merge_df<-merge(x=decrease_hw_df,y=infant_mortality_df,by= 1 ,all.x=TRUE)
```

```{r}
increase_merge_new<-increase_merge_df[increase_merge_df$Year.x == increase_merge_df$Year.y,]
decrease_merge_new<-decrease_merge_df[decrease_merge_df$Year.x == decrease_merge_df$Year.y,]
same_merge_new<-same_merge_df[same_merge_df$Year.x == same_merge_df$Year.y,]
```

```{r}
#Removing population column

same_merge_new<-same_merge_new[,-c(4)]
decrease_merge_new<-decrease_merge_new[,-c(4)]
increase_merge_new<-increase_merge_new[,-c(4)]

# Renamed Year column
same_merge_new %>% rename_at("Year.x", ~"year")

# Renamed Year column for the following df and added a new column for true or false
decrease_merge_new <- decrease_merge_new %>% rename_at("Year.x", ~"year") %>% mutate(increase_in_hygiene_facilities = FALSE)

# Renamed Year column for the following df and added a new column for true or false
increase_merge_new <- increase_merge_new %>% rename_at("Year.x", ~"year") %>% mutate(increase_in_hygiene_facilities = TRUE)

# Joined rows into one table
binded_unicef_df <- bind_rows(decrease_merge_new, increase_merge_new)

# Renamed columns so there's no spaces between words.
binded_unicef_df <- binded_unicef_df %>% rename_at("Infant Mortatlity per 1,000 Births", ~"death_rate_per_1000_births")

binded_unicef_df <- binded_unicef_df %>% rename_at("COUNTRY, AREA OR TERRITORY", ~"country")

# converted data type for year from chr to dbl
binded_unicef_df$year <- binded_unicef_df$year %>% as.numeric()

binded_unicef_df
```



```{r}
ggplot(binded_unicef_df, aes(year, death_rate_per_1000_births, color = increase_in_hygiene_facilities)) +
  geom_line() +
  facet_wrap(~country)
```

```{r}
yearly_summary <- binded_unicef_df %>%
  group_by(year, increase_in_hygiene_facilities) %>%
  summarise(mean_deaths = mean(death_rate_per_1000_births))

mean_deaths_by_boolean <- binded_unicef_df %>%
  group_by(increase_in_hygiene_facilities) %>%
  summarise(mean_deaths = mean(death_rate_per_1000_births))

mean_deaths_by_boolean
yearly_summary

ggplot(yearly_summary, aes(year, mean_deaths, color = increase_in_hygiene_facilities)) +
  geom_line()
```


```{r}

# Calculating a 95% Confidence interval using t.test 
test_result <- t.test( death_rate_per_1000_births ~ increase_in_hygiene_facilities, data = binded_unicef_df)
test_result

```







