---
title: "DATA607_Final_Project"
author: "Wilson Ng"
date: "2022-11-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mongolite)
library(tidyverse)
library(lubridate)
library(ggplot2)
library(keyring)
```

## 1.Connecting to MongoDB to access Dr.Semmelweis's data on childbirth deaths.

```{r}
# making rstudio ask me for the password since I don't want to expose my pw when I upload this file to GitHub.

mongo_server <- 'mongodb+srv://'
username <- 'wilsonnrdb:'
password <- .rs.askForPassword("Database Password:")
cluster <- '@cluster0.cyjo0wo.mongodb.net/'

connection_string = paste(mongo_server, username, password, cluster, sep = '')

# Two collections - one has yearly data and the other one has monthly data.
yearly_collection = mongo(collection='yearly', db='DATA607', url=connection_string)
monthly_collection = mongo(collection= 'monthly', db='DATA607', url=connection_string)
```

## 2.Examining the yearly data

```{r}
yearly_query = yearly_collection$find()

yearly_query
```
## 3.Examining the monthly data

```{r}
monthly_query = monthly_collection$find()

monthly_query
```
## 4.Calculating the porportion of deaths out of the number of women giving birth.

The table above shows the number of women giving birth at the two clinics at the Vienna General Hospital for the years 1841 to 1846.  

```{r}
# Adding a new column to yearly with proportion of deaths per no. births
yearly_query <- yearly_query %>% mutate(proportion_deaths = deaths / births)

# Converting Year column's data type to double.
yearly_query$year <- as.numeric(yearly_query$year)

# Print out yearly
yearly_query
```
## 5. Plotting the proportion of deaths at both clinic 1 and clinic 2.

```{r}
# Setting the size of plots in this notebook
options(repr.plot.width=7, repr.plot.height=4)

# Plot yearly proportion of deaths at the two clinics
ggplot(yearly_query, aes(year, proportion_deaths, color = clinic)) +
    geom_line()
```
## 6. Examining the monthly data from Clinic 1.

```{r}
# Adding a new column with proportion of deaths per no. births
monthly_query <- monthly_query %>% mutate(
    proportion_deaths = deaths / births
    )

# Converting Date column's chr data type into date data type
monthly_query$date <- ymd(monthly_query$date)

# Print out the first rows in monthly
head(monthly_query)
```
```{r}
# Plot monthly proportion of deaths to see if there's any obvious decline in deaths.
ggplot(monthly_query, aes(date, proportion_deaths)) +
    geom_line() +
    labs(x = "Date", y = "Proportion of Deaths", title = "Monthly Proportion of Deaths")
```
## 7. The effect of handwashing highlighted

Dr.Semmelweis made handwashing obligatory in the summer of 1847.
The effect of handwashing is made even more clear if we highlight this in the graph.

```{r}
# From this date handwashing was made mandatory
handwashing_start = as.Date('1847-06-01')

# Add a TRUE/FALSE column to monthly called handwashing_started
monthly_query <- monthly_query %>% mutate(
    handwashing_started = date >= handwashing_start
    )

# Plot monthly proportion of deaths before and after handwashing
ggplot(monthly_query, aes(date, proportion_deaths, color = handwashing_started)) +
    geom_line() +
    labs(x = "Date", y = "Proportion of Deaths", title = "Monthly Proportion of Deaths")
```
## 8.How much did it reduce the monthly proportion of deaths on average?

It reduced the proportion of deaths by around 8 percentage points.

```{r}
# Calculating the mean proportion of deaths 
# before and after handwashing.

monthly_summary <- monthly_query %>% 
    group_by(handwashing_started) %>%
    summarise(mean_proportion_deaths = mean(proportion_deaths))

# Printing out the summary.
monthly_summary
```
## A statistical analysis of Semmelweis handwashing data

Proportion of deaths increased by between 6.7 and 10 percentage points when doctors didn't wash their hands, according to a 95% confidence interval.

```{r}
# Calculating a 95% Confidence intrerval using t.test 
test_result <- t.test( proportion_deaths ~ handwashing_started, data = monthly_query)
test_result
```

